\subsection{Escapes}

The core of SERA will always be its transliteration definition for the Fidel
syllabary. SERA provides ``escapes'' or ``switches'' so that changes of
language and scripts can be signaled to a reader without requiring special
software to read the document. Special purpose escapes are also provided so
that applications may communicate graphic elements and processing directives in
an ASCII document. \\

%Software developers may wish to apply SERA's transliteration definition and
%forgo the escape specifications for their own proprietary system. This approach
%presents no complications when only the proprietary environment is used. The
%escape mechanism provided in SERA is then recommended when content is exported
%to simple text files. SERA transliteration applied in HTML documents, for
%example, is one such document type where an alternative escape system is
%available. \\

The backslash character then is chosen for escapes in SERA as it is in
agreement with the existing conventions of Unix, \LaTeX, C, and other
programming languages. 


\noi
\begin{tabular}{|l|l|} \hline
  $\backslash$   &   \parbox{2.0in}{\scriptsize Script Toggle ... $\leftarrow$ fidel $\rightarrow$ $\backslash$ $\leftarrow$ latin $\rightarrow$ $\backslash$ $\leftarrow$ fidel $\rightarrow$ etc. 
                     One or more punctuation characters following $\backslash$ will not 
                     require a closing $\backslash$  The script toggle terminates when the 
                     first nonpunctuation charcter is reached.}\\ \hline

  $\backslash\backslash$  &  \small{$\backslash$ The backslash escape is in follow-}\\ 
                &    \small{ing with normal rules for punctuation.}\\ \hline

  $\backslash\tilde{}\,$!   &   Escapes are ignored until closing $\backslash\tilde{}\,$!  \\ \hline



  $\backslash\tilde{}\,$xxx &  \\
  $\backslash\tilde{}$\{do xxx\}    &       Perform ``xxx''  \\ \hline
\end{tabular} \\

\noi
Whitespace is the required terminator for all escapes. When space, `` '',
is the following whitespace terminator it will be removed from the transcribed
output. 


\subsection{Multilingualism}


It is assumed that a document will be written primarily in two languages
-which may be written in one or two scripts. The regular or bilingual script 
escape, ``$\backslash$'' , always serves the two primary languages in the 
document. After switching to a third language, ``$\backslash$'' will indicate
a return to the first of the two major modes. \\

SERA applies the ISO 639 2 character and 3 character language names for
multilingualism. The principle is identical to that adopted in HTML 3.0. The 
language name is then simply appended to the special purpose escape 
``$\backslash\tilde{}$\,''. \\  

\noi
Example Usage: \\

\noi
\hspace*{-0.1in}
{\small
\begin{tabular}{lll}
    $\backslash\,\tilde{}$amh$\tilde{}$\,eng   
	                       & this is amharic   & (Set Primary/Secondary)    \\
    $\backslash\,\tilde{}$\,tir & this is tigrigna   & (New Third Language) \\
    $\backslash$           & this is amharic   & (Return To Primary)        \\
    $\backslash$           & this is english   & (Secondary)                \\
    $\backslash\,\tilde{}$ar\,$\tilde{}$gz     
	                       & this is arabic    & ({\footnotesize Reset Primary/Secondary})  \\
    $\backslash$           & this is ge'ez     & (Secondary)              
\end{tabular}   
}
